NumPy---Array Computation for Python

### 2020-01-04 body.tex
Minor edits
First pass - to give an idea of direction.

\subsection{History}

Before NumPy, two different Python array packages—Numeric and NumArray—existed.
The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
manipulation of many small arrays.
By the beginning of the millennium, several libraries built on top of Numeric 
to provide fundamental scientific routines were combined into SciPy.
Due to the poor performance of Numeric on large data sets, NumArray was
developed at the Space Telescope Science Institute (STScI) to handle
large astronomical images such as those produced by the Hubble Space Telescope.
However, while initially considered the successor to Numeric, NumArray was no longer
performant on multiple small arrays.
This divide posed a challenge for SciPy as it wasn't clear how to support
both libraries.
Working with the NumArray and SciPy communities, the then maintainer of Numeric,
Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.

The creation of NumPy spurred renewed development in both SciPy and the larger
scientific Python ecosystem and heralded the current era of wide-spread use of
Python for scientific computing.
To coordinate work as a global community, to ensure accuracy of computed
results, and to allow modification without breakage, the NumPy (and SciPy) teams
pioneered best software engineering practices in the scientific Python
community\cite{millman2014developing}.
For example, NumPy early on adopted distributed revision control, unit testing,
continuous integration, code review, and executable documentation.
It also developed a standard for ``docstrings''—text describing each NumPy function.
This quickly became a documentation standard for the larger scientific Python community.

Travis Oliphant eventually stepped back from leading NumPy to found a
company and to start new projects including Numba---an accelerated Python
compiler that could vastly accelerate NumPy expressions.  Development
of NumPy continued by a core group that informally operated by
consensus.

@@

\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
indexing and slicing}

>>>>

\subsection{Background}

% https://books.google.com/books?id=QQJODwAAQBAJ&pg=PA101&lpg=PA101&dq=%E2%80%9Cthe+secret+to+successful+hiring+is+this:+look+for+the+people+who+want+to+change+the+world.%E2%80%9D&source=bl&ots=D6l8JXfB20&sig=ACfU3U0x1X9Gh8k6MlNV0BqhBa_73Q0yfQ&hl=en&sa=X&ved=2ahUKEwj48pqg_oHgAhVohuAKHYlJAsc4ChDoATAAegQIChAB#v=onepage&q=%E2%80%9Cthe%20secret%20to%20successful%20hiring%20is%20this%3A%20look%20for%20the%20people%20who%20want%20to%20change%20the%20world.%E2%80%9D&f=false
% "The secret to successful hiring is this: look for the people who want to change the world.".  Marc Benioff.

NumPy is both a cause and an effect of a striking shift in the nature of
scientific computing.  The traditional model of scientific software has been of
the user as {\it consumer}.  Writing software is a task for specialists and
professionals, supported by income from selling the software to scientists to
use.  The new model is of the user as {\it owner}.  The Python language, with
the NumPy library, provides a framework that scientists can consume, but has
also been successful in encouraging scientists to develop as programmers, and
to write code for the library, and the system of packages around it.  This has
fostered a new culture of contribution and collaboration in scientific
computing that resulted in a virtuous cycle of increasing use, and increasing
development.

This cycle is one of several explanations for the astonishing growth of NumPy
in scientific computing.

The first incarnation of Numpy was a package called "Numeric", written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Huginin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  At the time, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large numbers of full-time professional programmers
and supported by formal quality assurance, release managers and writers of
scientific documentation.  In contrast, Numeric, and later, Numpy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding, and often to the detriment of
the scientists' careers.

Table X1 about here: backgrounds of initial designers (mainly math, physics).

Table X2 about here: backgrounds of major committers (mainly math, physics).

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a re-implementation of much of
Numeric, called Numarray, to support their need to work on large memory-mapped
arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and Numarray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.

Over the next 10 years, Numpy attracted a relatively large pool of scientist
developers (see Figure ~\ref{fig:prs-over-time}).  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improved collaboration on code, and
continuous testing so we run a large battery automated tests for every proposed
change in Numpy. At first, Numpy had rather sparse documentation, but the project
developed a new standard for ``docstrings'' —- text describing each NumPy
function -- and used this standard to write comprehensive high-quality
documention, integrated with the source code. The Numpy documentation format
quickly became a standard for the larger Python community.

As we implied above, there were several factors that allowed this rapid growth
and successful development.

Hugunin identified the first factor in his initial description of the Numeric
package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
component of a larger programming task.  Therefore:

\begin{quote}
    Rather than trying to retrofit an existing numerical language to support
    the wealth of features found in a powerful, modern, general-purpose
    programming language, it makes much more sense to attack the problem from
    the other direction and add the features of a powerful numerical
    programming language to Python.
\end{quote}

Python is already a good choice for many standard programming tasks such as
cleaning data, interacting with web resources and parsing text.  It has a wide
range of libraries for many different tasks. Adding fast array operations and
linear algebra allows the scientist to do all their work with within the same
language - and one that has the advantage of being famously easy to learn and
teach.

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, Numpy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development.  In contrast, Numpy often
attracts user-developers, who want to work with Numpy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing Numpy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

The third factor is much like the second: a culture in Python that concentrates
on high quality of code and distribution. Because Python is open-source, it has
a strong culture of contribution, and therefore, of using tools and process to
improve collaborative software development, such as distributed version
control, code review, and automated testing.  These have allowed the project to
scale safely as it attracted more users and developers, even though these
developers rarely came to the project with training in software engineering.

Lastly, we believe Numpy has benefitted greatly from a well-chosen application
programming interface (API).  Scientific and numerical programming needs to be
fast, and scale to very large datasets.  The Numpy API defines a simple wrapper
for data in memory so it can be represented as a one- or multi-dimensional
array.  The simplicity of this wrapper has made it very successful as a
standard way of representing arrays in memory, and has made it relatively easy
for other libraries to develop fast and memory-efficient compiled code, usually
in C or Fortran, that can manipulate these arrays and pass them back to Python.
This has been a large factor in the growth of numerical computing libaries
around Numpy and Python.

@@

\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
indexing and slicing, one-liners that would be tens or hundreds of lines of code in
languages such as C.}
